NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
Edits to the counterion binding section

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13} whereas
the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting order parameter values
are shown in the supplementary information (Fig. \ref{DPslices}).

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences substantial forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
%Moved here from the molecular electrometer paragraph because there it distrupted the flow of the section and it contains results. This also enables referring to figures in order. I also removed the refs which do not seem to refer to the actual data used in the figure.
%Rewrote the paragraph, see if ok.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
    and MD simulations with different force fields (symbols).
    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data are without additional salt.
    The data for DOPS is at 303~K and the data for POPS is at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the symbol marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations around
    (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
    of PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at~\citenum{POPCcharmm36T303K}.
  }

\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
    Rough, subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
    Here M indicates a magnitude problem, F a forking problem; letter size increases with problem severity. Color scheme: within experimental error (dark green), almost within experimental errorÓ (light green), Òclear deviation from experiments (light red), and major deviation from experiments (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see SI section~\ref{Ranking}.
  }
\end{figure}

The different PS MD models produce highly varied  headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
as previously observed also for PC lipids \cite{botan15}, and none of the models produces a set of order parameters in full agreement with the experiments.
The models perform generally less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
which complicates the interpretation of structural differences between PC and PS headgroups. However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
also observed in the experiments (Fig. \ref{HGorderParameters}).

@@

Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
dihedral angle (single peak around 120$^{\circ}$) in comparison to the other models which yield a distribution between two angles (Fig. \ref{dihedralsHG}).
The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
and narrow angle distribution for PS than for PC headgroup in
CHARMM36 simulations that have the best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
% Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),but further analysis is beyond the scope of this work which focuses on the PS headgroup. There is very little actual information in this statement. The variance was already noted above

These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments. However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to confirm the correct conformational ensemble.

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSvsPC}
    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
    Experimental values with the signs are measured for pure POPS system in this work.
    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
    Because the experimental values of POPS in pure and mixed bilayers come from $^{13}$C NMR (this work) and $^2$H NMR (Ref. \citenum{roux90}), respectively,
    the error bars of 0.02 are used here \cite{botan15,ollila16}.
    The y-axis for the $\alpha$-carbon results of POPS (bottom) is transferred
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
    The values from CHARMM36 and Gromos simulations are averages of two independent simulations and
    the error bars are given as the difference between the results divided by two.
  }
\end{figure}

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. 
Counterion density profiles along membrane normal 
show significant differences between force fields in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in Gromos-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller areas (Fig. \ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values.
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36ua simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
In conclusion, the results are in line with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.

@@

The experimentally observed modulation of headgroup order parameters
by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions and the lack of a ion-free reference state.
The analysis is further complicated by the artificial aggregation of counterions
observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)
% \todo{Why isn't this a problem with PC}.
% SAMULI: The difference between ion-free state (present in PC) and low amount of monovalent ions is larger than
% between counterion only and low amount of monovalent ions in charged lipids. I believe that therefore we can
% make conclusions from smaller ion concentrations without aggregation in PC lipids, but not in charged lipids.
Therefore, we evaluate here the amount of bound charge not by adding salt
(although also this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
but by studying the changes of the headgroup order parameters with increasing amount of
negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.

According to the molecular electrometer concept, the headgroup order parameters of POPC
increase when negatively charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36 and Berger simulations either exhibit no change or show a decrease
in the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
This can be explained by the stronger counterion binding affinity, which cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
realistic surface charge density in membranes containing PS lipids,
while the other tested simulation models overestimate the counterion
binding affinity. The results are in line with the behaviour of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the SI.

The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
together with other experimental results suggest less rigid structure of PS headgroups when diluted with
POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
and we conclude that more accurate force fields are needed
to correctly describe the PC-PS headgroup interactions in MD simulations.
%Having this conclusion in every paragraph gets a bit repetive

>>>>

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting order parameter values
are shown in the supplementary information (Fig. \ref{DPslices}).

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
%Moved here from the molecular electrometer paragraph because there it distrupted the flow of the section and it contains results. This also enables referring to figures in order. I also removed the refs which do not seem to refer to the actual data used in the figure.
%Rewrote the paragraph, see if ok.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
    and MD simulations with different force fields (symbols).
    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data are without additional salt.
    The data for DOPS is at 303~K and the data for POPS is at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the dot marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from simulations conducted with CHARMM36 --- the force field producing the best agreement with experiments ---
     demonstrate the conformational fluctuations around
    (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
    of PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of PS and PC headgroups. The publicly available trajectory was used for CHARMM36 POPC~\cite{POPCcharmm36T303K}.
  }
%  \todo{Figure clarity is questioned in: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/48}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
    Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
    Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity. Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
\end{figure}

The different PS MD models produce a wide variety of headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
as previously observed also for PC lipids \cite{botan15}. None of the models produces a set of order parameters in full agreement with the experiments.
The models perform generally less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
which complicates the interpretation of structural differences between PC and PS headgroups.

@@

However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
also observed in the experiments (Fig. \ref{HGorderParameters}).

Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
dihedral angle (single peak around 120$^{\circ}$) in comparison to the other models which yield a distribution between two angles (Fig. \ref{dihedralsHG}).
The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
and narrow angle distribution for PS than for PC headgroup in
CHARMM36 simulations that have the best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
% Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),but further analysis is beyond the scope of this work which focuses on the PS headgroup. There is very little actual information in this statement. The variance was already noted above

The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments. However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to confirm the correct conformational ensemble.

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSvsPC}
    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
    Experimental values with the signs are measured for pure POPS system in this work.
    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
    Because the experimental values of POPS in pure and mixed bilayers come from $^{13}$C NMR (this work) and $^2$H NMR (Ref. \citenum{roux90}), respectively,
    the error bars of 0.02 are used here \cite{botan15,ollila16}.
    The y-axis for the $\alpha$-carbon results of POPS (bottom) is shifted
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
    The values from CHARMM36 and Gromos simulations are averages of two independent simulations and
    the error bars are given as the difference between the results divided by two.
  }
\end{figure}

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. We explore this by quantifying the counterion density profiles along the membrane normal accompanied by the areas per lipid in Fig.~\ref{NAdensPOPS}.
The force fields studied show significant differences in both binding affinity
and distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in Gromos-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller values.
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36ua simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are in line with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.

@@

The experimentally observed modulation of headgroup order parameters
by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions (the lack of a ion-free reference state) and the analysis is further complicated by the artificial aggregation of counterions
observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)
% \todo{Why isn't this a problem with PC}.
% SAMULI: The difference between ion-free state (present in PC) and low amount of monovalent ions is larger than
% between counterion only and low amount of monovalent ions in charged lipids. I believe that therefore we can
% make conclusions from smaller ion concentrations without aggregation in PC lipids, but not in charged lipids.
Therefore, we evaluate here the amount of bound charge not by adding salt
(although this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
but by studying the changes of the headgroup order parameters with increasing amount of
negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.

Experimentally, the headgroup order parameters of POPC
increase when negatively charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36 and Berger simulations either exhibit no change or show a decrease
in the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
Therein, the stronger counterion binding cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
realistic surface charge density in membranes containing PS lipids,
while the other tested models overestimate the counterion
binding affinity. The results are in line with the behaviour of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the SI.

The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
together with other experimental results suggest that the PS headgroup structure becomes less rigid when diluted with
POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC})
and more accurate force fields are needed
to correctly describe the PC-PS headgroup interactions in MD simulations.
%Having this conclusion in every paragraph gets a bit repetive
